Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2023-05-22 write-up/fda-recurrent.tex
updated main and supp papers

\title[Recurrent event analysis with functional covariates via random subsampling]{Recurrent event analysis in the presence of real-time high frequency data via random subsampling}
% \author{Walter Dempsey}
% \address {Department of Biostatistics, University of Michigan, 1415 Washington Heights, Ann Arbor, MI 48109, USA}
% \email{wdem@umich.edu}

@@

In this paper, we also consider joint modeling of recurrent events and a high frequency sensor process. In contrast to traditional longitudinal biomarkers, sensor processes are observed multiple times per second leading to thousand and/or millions of observations per individual.  To address this increase in scale, a design-based perspective is taken in which we subsample non-event times to avoid some of the computationally demanding aspects of joint models.  Subsampling in the context of functional data analysis for massive data has recently been investigated~\citep{https://doi.org/10.48550/arxiv.2104.03446} in which a functional L-optimality criterion is derived. Subsampling is different in our current context, as our goal is to subsample from the individual sensor processes to produce a design-unbiased estimate of the cumulative hazard function.  Finally, traditional joint modeling often consider the hazard function to depend on the current or cumulative value of the health process~\citep{Li2022,Rizopoulos2010}. For high frequency sensor processes, we expect the hazard function to only depend on the recent sensor history and use recent history functional linear models~\cite{KIM20111554} to account for this local dependence.  We combine these models with advances in longitudinal functional data analysis~\citep{Staicu2010} and penalized functional regression~\citep{Goldsmith2015} to perform scalable inference using off-the-shelf software.

@@

Suppose $n$ subjects are independently sampled with observed event times~$\bfT_{i} = \{ T_{i,1}, \ldots, T_{i,k_i}\}$ over some observation window $[0, \tau_i]$ for each subject $i = 1,\ldots, n$.  Assume the event times are ordered, i.e., $T_{i,j} < T_{i,j^\prime}$ for $j < j^\prime$. The observation window length, $\tau_i$, is the censoring time and is assumed independent of the event process. Let~$N_{i} (t)$ denote the associated counting process of $\bfT_{i}$; that is, $N_i (t) = \sum_{j=1}^{k_i} 1 [ T_{i,j} < t ]$. In this section, we assume a single-dimensional health process~$\bfx_i = \{ x_i (s) \}_{0 < s < \tau_i}$ for each participant is measured at a dense grid of time points.  Accelerometer, for example, is measured at a rate of 32Hz (i.e., $32$ times per second). Electrodermal activity (EDA), on the other hand, is measured at a rate of 4Hz (i.e., 4 times per second).  Given the high frequency nature of sensor data, this paper assumes the process is measured continuously.

@@

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. The component of the log-likelihood related to the event process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right),
\]
with $h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

@@

\begin{rmk}[Connection to design-based estimation]
The approximate score equations given by~\eqref{eq:approxscore} arise in design-based inference of point processes.  Design-based inference is common for spatial point processes~\citep{Waagepetersen2008} where the spatial varying covariate is observed at a random sample of locations. It is common in mobile health where ecological momentary assessments~\citep{Rathbun2012,Rathbun2016} are used to randomly sample individuals at various time-points to assess their emotional state.  In the current setting, we leverage these ideas to form a subsampling protocol that can substantially reduce computationally complexity.  Therefore, the purpose is quite different. Moreover, the dependence of the intensity function on the recent history of sensor values leads to additional complications that must be addressed.
\end{rmk}

@@

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,\textcolor{blue}{\tau_i}]$ denote the random set of subsampled points.  Note, by construction, this random set is distinct from the set of event times with probabilty one, i.e.,~$\pr( \bfT_i \cap \bfD_i = \emptyset) = 1$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard $\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} h^{(1)}_i (u; \theta)/\pi_i (u)$. An alternative design-unbiased estimator of the derivative of the cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Proposition~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting approximate estimating equations can be re-written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)}_i (u; \theta)}{ h_i ( u; \theta)}  - \sum_{u \in \bfD_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} represents the approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

@@

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (T,s), (T,s^\prime) ) f_y(T) dT.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((t,s), (t,s^\prime))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$, $\tau$ is the observation window length of the process, and $f_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (t,s_r) \tilde X (t,s_{r^\prime}) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

@@

\begin{rmk}[Inference procedure review] \normalfont
Given observed recurrent event and high frequency data~$\{ \bfT_i, \bfx_i \}_{i=1}^n$,
\begin{enumerate}[(a)]
\item For each $i \in [n]$, sample non-event times as a time-inhomogeneous Poisson point process with intensity according to~$\pi_i (t)$
\item \label{p2} Estimate mean~$\mu_y (t,s)$ for $0 \leq s \leq \Delta$ at all event times~$t \in \cup_{i=1}^n \bfT_i$ and sampled non-event times~$t \in \cup_{i=1}^n \bfD_i$.
\item \label{p3} Compute marginal covariance across event times,~$\Sigma_1$, and non-event times,~$\Sigma_0$.
\item \label{p4} Compute eigendecomposition~$\{ \hat \psi_k^{(y)}, \hat \lambda_k^{()} \}$ of marginal covariance~$\Sigma_y$
\item Use the eigendecomposition to construct~$W_{i,t}$ for all $i \in [n]$ and $t \in \bfD_i \cup \bfT_i$
\item \label{point:log} Perform logistic regression with binary outcome~$\{ \Y_i (t) \}$  and offset of~$\log \pi_i (t)$.
\end{enumerate}
\end{rmk}

@@

Our theoretical analysis requires assumptions regarding the subsampling procedure, the event process, and the functional data. We state these assumptions and then our main theorems. We start by assuming there exists a~$\tau < \infty$ such that all individuals are no longer at risk (i.e.,~$\tau_i < \tau$ for all~$i$). Moreover, define $R_i (t)$ to be the at-risk indicator for participant~$i$, i.e., $R_i (t) = {\bf 1} [t \in (0,\tau_i)]$. Asymptotic theory provided in Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following additional assumptions:

@@

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(t,s)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(t, s) : (t,s) \in \mathcal{T} \times
  \mathcal{S} \}$ is a square integrable element of the $L^2 (
  \mathcal{T} \times \mathcal{S})$
\item\label{A2} The subsampling and conditional intensity rate functions~$f_y(T)$ are continuous and~$\sup |f_y(T)| < \infty$.
\item\label{A3} $\E[X(t,s) X(t,s^\prime) X(t^\prime,s) X(t^\prime, s^\prime) ] < \infty$ for each $s,s^\prime \in [0,\Delta]$ and~$0 < t, t^\prime < \tau$.
\item\label{A4} $\E[\|X(t,\cdot)\|^4] < \infty$ for each~$0< t < \tau$.
\end{enumerate}
\end{assumption}
Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion.
\begin{rmk}[Practical consequences of Assumptions~\ref{assumption:events} and~\ref{assumption:truncation}]
Bounded and iid.
\end{rmk}

With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.

\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events}, Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \times \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
A design-unbiased estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$.
\begin{newmaterial}
Based on this, we can leverage existing inferential machinery to obtain variance-covariance estimates of model parameters. That is, if~$\hat \Sigma_{bb}$ is the $K_b \times K_b$ dimensional matrix obtained by plugging in the estimates of variance components into the formula for the variance of~$\hat b$, then the standard error for estimate at time~$t_0$ -- i.e., ~$\hat \beta (t_0) = \phi(t_0) \hat {\bf b}$ -- is given by~$\sqrt{ \phi (t_0 ) \hat \Sigma_{bb} \phi(t_0)^\top}$.  Then the approximate 95\% confidence interval can be constructed as~$\hat \beta (t_0) \pm 1.96 \sqrt{\phi (t_0) \hat \Sigma_{bb} \phi(t_0)^\top}$.  We acknowledge an important limitations of confidence intervals obtained via this approach. Specifically, we ignore the variability inherent in the longitudinal functional principal component analysis; that is, our estimates ignore the variability in estimation of eigenfunctions~$\hat \psi$ as well as the coefficients~$\hat c_{i,k}(t)$. Joint modeling could be considered as in~\cite{Crainiceanu2010}, however, this is beyond the scope of this article.

Lemma~\ref{prop:optimal} shows~\eqref{eq:waage_weights} is optimal within a class of weighted estimating equations. The result ensures the only loss of statistical efficiency is due to subsampling and not using a suboptimal estimation procedure given subsampling.  Here, weights $w^\star$ are considered \emph{optimal} if the difference between the asympotic variance~$V (\theta_0; w^\star)$ and the asymptotic variance under any other choice of weights~$w$, $V(\theta_0; w)$ is positive semi-definite, i.e., any linear contrast has smaller asymptotic variance under weight $w^\star$ than under weight~$w$.
\end{newmaterial}
\begin{lemma} \normalfont
\label{prop:optimal}
% \textcolor{red}{
If the event process is an inhomogeneous Poisson point process with intensity~$h(t; \theta)$ and subsampling occurs via an independent, inhomogeneous Poisson point process with intensity~$\pi (t)$, then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most efficient) in the class of weighted estimating functions given by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any weight function~$w_i (t; \theta)$. This class includes the Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
% WD: Consider larger class and show the proof doesn't extend to non PP models?}
\end{lemma}

\noindent

@@

Under assumption~\ref{assumption:truncation}, we next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. In particular, we consider the following choice of subsampling rate,~$\pi(t) = c \times h(t; \theta)$ for $c>0$. That is, the subsampling rate is proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c/ (c+1)$. Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H = \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

@@

\subsection{Computational comparison to alternative methods} \label{section:simcomparisons}

@@

Inspection of Figures~\ref{fig:mean_eda} and~\ref{fig:mean_acc} suggest setting $\Delta = 30$ minutes is adequate to capturing the proximal impact of EDA and AI on the risk of a button press.  To ensure minimal loss of efficiency, the subsampling rate was set to once every fifteen minutes.  Given the daily button press rate, this ensures an average of $44$ non-events to $2.5$ events per day.  Based on Table~\ref{tab:compvseff}, this ensures we can achieve a substantial data reduction at a minimal loss of efficiency.  After sampling non-event times, complete-case analyses are performed, i.e., sampled times where sensors included in the model have any level of missing data are ignored.

>>>>

\title[Recurrent event analysis with functional covariates via random subsampling]{Recurrent event analysis in the presence of real-time high frequency data via random subsampling}
\author{Walter Dempsey}
\address {Department of Biostatistics, University of Michigan, 1415 Washington Heights, Ann Arbor, MI 48109, USA}
\email{wdem@umich.edu}

@@

In this paper, we also consider joint modeling of recurrent events and a high frequency sensor process. In contrast to traditional longitudinal biomarkers, sensor processes are observed multiple times per second leading to thousand and/or millions of observations per individual.  To address this increase in scale, a design-based perspective is taken in which we subsample non-event times to avoid some of the computationally demanding aspects of joint models.  Subsampling in the context of functional data analysis for massive data has recently been investigated~\citep{https://doi.org/10.48550/arxiv.2104.03446} in which a functional L-optimality criterion is derived. Subsampling is different in our current context, as our goal is to subsample from the individual sensor processes to produce a design-unbiased estimate of the cumulative hazard function.  Finally, traditional joint modeling often consider the hazard function to depend on the current or cumulative value of the health process~\citep{Li2022,Rizopoulos2010}. For high frequency sensor processes, we expect the hazard function to only depend on the recent sensor history and use recent history functional linear models~\citep{KIM20111554} to account for this local dependence.  We combine these models with advances in longitudinal functional data analysis~\citep{Staicu2010} and penalized functional regression~\citep{Goldsmith2015} to perform scalable inference using off-the-shelf software.

@@

Suppose $n$ subjects are independently sampled with observed event times~$\bfT_{i} = \{ T_{i,1}, \ldots, T_{i,k_i}\}$ over some observation window $[0, \tau_i]$ for each subject $i = 1,\ldots, n$.  Assume the event times are ordered, i.e., $T_{i,j} < T_{i,j^\prime}$ for $j < j^\prime$. The observation window length, $\tau_i$, is the censoring time and is assumed independent of the event process. Let~$N_{i} (t)$ denote the associated counting process of $\bfT_{i}$; that is, $N_i (t) = \sum_{j=1}^{k_i} 1 [ T_{i,j} < t ]$. In this section, we assume a single-dimensional health process~$\bfx_i = \{ x_i (s) \}_{0 \le s < \tau_i}$ for each participant is measured at a dense grid of time points.  Accelerometer, for example, is measured at a rate of 32Hz (i.e., $32$ times per second). Electrodermal activity (EDA), on the other hand, is measured at a rate of 4Hz (i.e., 4 times per second).  Given the high frequency nature of sensor data, this paper assumes the process is measured continuously. \textcolor{blue}{See Appendix~\ref{app:notation} for a notation glossary.}

@@

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. \textcolor{blue}{In our current setting, we assume the health process $\{ x_i(s)\}_{0 \leq s < \tau_i}$ is directly observed.  Therefore, we can consider the event process~$\bfT_i$ conditional on the health process~$\bfx_i$, which results in the log-likelihood related to the event process being then given by}
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. \textcolor{blue}{See Appendix~\ref{section:memproblems} for additional arguments in favor of our proposed approach.} Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right),
\]
with $h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

@@

% \begin{rmk}[Design-based estimation]
\textcolor{blue}{
To circumvent these computational difficulties, we will derive approximate score equations based on design-based inference of point processes.  Design-based inference is common for spatial point processes~\citep{Waagepetersen2008} where the spatial varying covariate is observed at a random sample of locations. It is common in mobile health where ecological momentary assessments~\citep{Rathbun2012,Rathbun2016} are used to randomly sample individuals at various time-points to assess their emotional state.  In the current setting, we will leverage these ideas to form a subsampling protocol that can substantially reduce computationally complexity.  Therefore, the purpose is quite different. Moreover, the dependence of the intensity function on the recent history of sensor values leads to additional complications that must be addressed.}
% \end{rmk}

@@

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,\textcolor{blue}{\tau_i}]$ denote the random set of subsampled points.  Note, by construction, this random set is distinct from the set of event times with probabilty one, i.e.,~$\pr( \bfT_i \cap \bfD_i = \emptyset) = 1$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard $\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} h^{(1)}_i (u; \theta)/\pi_i (u)$. An alternative design-unbiased estimator of the derivative of the cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Lemma~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting approximate estimating equations can be re-written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)}_i (u; \theta)}{ h_i ( u; \theta)}  - \sum_{u \in \bfD_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} represents the approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

@@

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (T,s), (T,s^\prime) ) f_y(T) dT.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((t,s), (t,s^\prime))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$, $\tau$ is the observation window length of the process, and $f_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (t,s_r) \tilde X (t,s_{r^\prime}) \right).
\]
\textcolor{blue}{As we assume the health process~$\bfx_i$ is directly observed}, the diagonal elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

@@

\begin{rmk}[Inference procedure review] \normalfont
\label{rmk:review}
Given observed recurrent event and high frequency data~$\{ \bfT_i, \bfx_i \}_{i=1}^n$,
\begin{enumerate}[(a)]
\item For each $i \in [n]$, sample non-event times as a time-inhomogeneous Poisson point process with intensity according to~$\pi_i (t)$
\item \label{p2} Estimate mean~$\mu_y (t,s)$ for $0 \leq s \leq \Delta$ at all event times~$t \in \cup_{i=1}^n \bfT_i$ and sampled non-event times~$t \in \cup_{i=1}^n \bfD_i$.
\item \label{p3} Compute marginal covariance across event times,~$\Sigma_1$, and non-event times,~$\Sigma_0$.
\item \label{p4} Compute eigendecomposition~$\{ \hat \psi_k^{(y)}, \hat \lambda_k^{(y)} \}$ of marginal covariance~$\Sigma_y$
\item Use the eigendecomposition to construct~$W_{i,t}$ for all $i \in [n]$ and $t \in \bfD_i \cup \bfT_i$
\item \label{point:log} Perform logistic regression with binary outcome~$\{ \Y_i (t) \}$  and offset of~$\log \pi_i (t)$.
\end{enumerate}
\end{rmk}

@@

Our theoretical analysis requires assumptions regarding the subsampling procedure, the event process, and the functional data. We state these assumptions and then our main theorems. We start by assuming there exists~$\tau < \infty$ such that all individuals are no longer at risk (i.e.,~$\tau_i < \tau$ for all~$i$). Moreover, define $R_i (t)$ to be the at-risk indicator for participant~$i$, i.e., $R_i (t) = {\bf 1} [t \in (0,\tau_i)]$. Asymptotic theory provided in Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following additional assumptions:

@@

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(t,s)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(t, s) : (t,s) \in \mathcal{T} \times
  \mathcal{S} \}$ is a square integrable element of the $L^2 (
  \mathcal{T} \times \mathcal{S})$
\item\label{A2} The subsampling and conditional intensity rate functions~$f_y(T)$ are continuous and~$\sup |f_y(T)| < \infty$.
\item\label{A3} $\E[X(t,s) X(t,s^\prime) X(t^\prime,s) X(t^\prime, s^\prime) ] < \infty$ for each $s,s^\prime \in [0,\Delta]$ and~$0 < t, t^\prime < \tau$.
\item\label{A4} $\E[\|X(t,\cdot)\|^4] < \infty$ for each~$0< t < \tau$.
\end{enumerate}
\end{assumption}
Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion.

\begin{newmaterial}
\begin{rmk}[Practical consequences of Assumptions~\ref{assumption:events} and~\ref{assumption:truncation}]
Assumptions~\ref{assumption:events} and~\ref{assumption:truncation} contain as a special case the scenario where individuals are independent and identically distributed, the functional process is bounded (i.e., $|X(t,s)| < M$ for some $M <\infty$), and the subsampling rate is both lower and upper bounded at all risk times. As such bounds are likely to be true in most practical settings, this demonstrates the reasonableness of our assumptions for applied settings.
\end{rmk}
\end{newmaterial}

% With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events}, Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \overset{D}{\to} N(0, \Xi (\theta)^{-1}),
\]
where $\overset{D}{\to}$ is convergence is distribution and
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \times \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
\textcolor{blue}{Proof of Lemma~\ref{lemma:simpleasym} is presented in Appendix~\ref{proof:asymp}.} A design-unbiased estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$.
\begin{newmaterial}
Based on this, we can leverage existing inferential machinery to obtain variance-covariance estimates of model parameters. That is, if~$\hat \Sigma_{bb}$ is the $K_b \times K_b$ dimensional matrix obtained by plugging in the estimates of variance components into the formula for the variance of~$\hat b$, then the standard error for estimate at time~$t_0$ -- i.e., ~$\hat \beta (t_0) = \phi(t_0) \hat {\bf b}$ -- is given by~$\sqrt{ \phi (t_0 ) \hat \Sigma_{bb} \phi(t_0)^\top}$.  Then the approximate 95\% confidence interval can be constructed as~$\hat \beta (t_0) \pm 1.96 \sqrt{\phi (t_0) \hat \Sigma_{bb} \phi(t_0)^\top}$.  We acknowledge an important limitations of confidence intervals obtained via this approach. Specifically, we ignore the variability inherent in the longitudinal functional principal component analysis; that is, our estimates ignore the variability in estimation of eigenfunctions~$\hat \psi$ as well as the coefficients~$\hat c_{i,k}(t)$. Joint modeling could be considered as in~\cite{Crainiceanu2010}, however, this is beyond the scope of this article.

Lemma~\ref{prop:optimal} shows~\eqref{eq:waage_weights} is optimal within a class of weighted estimating equations. The result ensures the only loss of statistical efficiency is due to subsampling and not using a suboptimal estimation procedure given subsampling.  Here, weights $w^\star$ are considered \emph{optimal} if the difference between the asympotic variance~$V (\theta_0; w^\star)$ and the asymptotic variance under any other choice of weights~$w$, $V(\theta_0; w)$ is positive semi-definite, i.e., any linear contrast has smaller asymptotic variance under weight $w^\star$ than under weight~$w$.
\end{newmaterial}
\begin{lemma} \normalfont
\label{prop:optimal}
% \textcolor{red}{
If the event process is an inhomogeneous Poisson point process with intensity~$h(t; \theta)$ and subsampling occurs via an independent, inhomogeneous Poisson point process with intensity~$\pi (t)$, then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most efficient) in the class of weighted estimating functions given by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any weight function~$w_i (t; \theta)$. This class includes the Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
% WD: Consider larger class and show the proof doesn't extend to non PP models?}
\end{lemma}
\noindent \textcolor{blue}{Proof of Lemma~\ref{prop:optimal} is presented in Appendix~\ref{app:optimalweights}.}

@@

We next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. In particular, we consider the following choice of subsampling rate,~$\pi(t) = c \times h(t; \theta)$ for $c>0$. That is, the subsampling rate is proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c/ (c+1)$. Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H = \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

@@

% \subsection{Computational comparison to alternative methods} \label{section:simcomparisons}

@@

\begin{newmaterial}
